---
title: "Super Tricky Challenge"
author: "Tim Riffe"
date: "July 8, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A challenging worked example

The challenge: how much can fertility distributions vary within a particular **level** of fertility? There's more than one way to do this. The example is challenging because the input data, downloaded as-is from the [Human Fertility Collection](https://www.fertilitydata.org/cgi-bin/index.php) have many overlapping subsets, such as overlapping periods, age-definitions, or even multiple sources for the same data. In principle we don't care about the specifics of each subpopulation, because we're interested in seeing a big pattern emerge from the data.

```{r, fig.height = 10, fig.width = 5}
library(tidyverse)

HFC <- read_csv("Data/HFC_ASFRstand_TOT.zip", 
                na = '.') 

HFCextremes <- 
  HFC %>% 
  # there's no variable for period width in the data, but we
  # need to be able to group on this too!! I found out the hard way...
  mutate(YearInt = Year2 - Year1 + 1) %>% 
  # I don't know if we really need to group on all this, but it seems
  # conservative enough to split on everything that hypothetically could
  # be used to define a subset
	group_by(Country, 
          Region,
          Urban,
          Origin,
					Year1, 
					AgeDef, 
					Vitality, 
					Collection, 
					SourceType, 
					RefCode,
					YearInt)  %>% 
  # This part narrated separately
	mutate(
	     Age = if_else(AgeDef == "ACY", Age + .5, Age),
	     TFR = sum(ASFR, na.rm = TRUE),
		   MAB = sum(ASFR * Age, na.rm = TRUE) / TFR,
		   TFRint = TFR - TFR %% .25) %>% 
	ungroup() %>% 
  # define extremes within TFR bins, then filter 
  # down to just these
	group_by(TFRint) %>% 
	mutate(MABmax = MAB == max(MAB),
		     MABmin = MAB == min(MAB)) %>% 
	filter((MABmax | MABmin) & (TFR < 6)) %>% 
	mutate(extremes = if_else(MABmax,"high MAB", "low MAB"))

library(ggridges)
library(colorspace)
HFCextremes %>% 
  ggplot(mapping = aes(x = Age, 
                       # this `y` sets the levels
			   					     y = as.factor(TFRint))) +
  # new geom!
	geom_ridgeline(aes(height = ASFR, # height is for curves within the levels
	                   fill = extremes), 
				   scale = 4, alpha = .5) +
  # clean look
  theme_minimal() + 
  labs(y = "asfr curves by TFR",
       title = "Many ways to have low fertility\none way to have high fertility",
       caption = "Data from HFC: www.fertilitydata.org") +
  theme(legend.position = "none") +
  # see hcl_palettes(plot=TRUE) to choose a different palette
  scale_fill_discrete_qualitative("Harmonic")
  # type `color picker` in google to get a quick way to find these values
  # scale_fill_manual(values = c("high MAB" = "#6384ba",
  #                              "low MAB" = "#8970c4"))
```

Separate annotation of the middle mutate, which looks like this:
```{r, eval = FALSE}
	mutate(
	     Age = if_else(AgeDef == "ACY", Age + .5, Age),
	     TFR = sum(ASFR, na.rm = TRUE),
		   MAB = sum(ASFR * Age, na.rm = TRUE) / TFR,
		   TFRint = TFR - TFR %% .25)
```
1 . We shift some `Age` classes up by 1/2 due to different **Lexis** shapes. If age is defined the way we usually think about it: "what was your last birthday?" (code = `ACY`), then we add 1/2 in order to center the rates in the interval. If age is defined as "what birthday gets celebrated this calendar year?" (this is the one in the CED logo, code  = `ARDY`), then it's already centered on an integer age. 
2. `TFR` is the sum of the rates over age. Here, this works for all groups, because we declared strict groups in the previous step.
3. `MAB` is the weighted average of `Age`, where age-specific fertility rates are the weights! The `TFR` you see in the denominator is just $\sum F_x$
4. We then make a variable for arbitrary TFR intervals using the modulo operator. 

* recall that `TFR`, `MAB`, and `TFRint` are summary indices, but we did them in `mutate()`, so they will be repeated over rows within groups!

The final bit of that pipeline is this:
```{r, eval = FALSE}
	group_by(TFRint) %>% 
	mutate(MABmax = MAB == max(MAB),
		     MABmin = MAB == min(MAB)) %>% 
	filter((MABmax | MABmin) & (TFR < 6)) %>% 
	mutate(extremes = if_else(MABmax,"high MAB", "low MAB"))
```
Here we create `TRUE` /  `FALSE` indicators for which subsets are the maximum and minimum mean ages at birth (`MAB`) *within* each `TFR` interval. We then filter down to only keep these extreme cases. Finally, we label the high and low subsets to be able to differentiate them in the plot.


I wonder how we might quantify how different two ASFR patterns of the same level are? Maybe convert to densities and see what fraction overlaps, like so:

```{r}
HFCextremes %>% 
  # select just what we need
  select(Age, ASFR, TFRint, extremes) %>% 
  # make sure no NAs; make age on par again
  mutate(ASFR = if_else(is.na(ASFR),0, ASFR),
         Age = floor(Age)) %>% 
  # standardize asfr to sum to 1, so we just have "shape"
  group_by(TFRint, extremes) %>% 
  mutate(asfr_stand = ASFR / sum(ASFR, na.rm = TRUE)) %>% # sum to 1
  ungroup() %>% 
  # put extreme asfr_stand side-by-side
  select(-ASFR) %>% 
  pivot_wider(names_from = extremes, values_from = asfr_stand) %>% 
  # again, this may have introduced NAs (not sure), due to different age ranges?
  mutate(`high MAB` = if_else(is.na(`high MAB`), 0,`high MAB`),
         `low MAB` = if_else(is.na(`low MAB`), 0,`low MAB`)) %>% 
  group_by(TFRint) %>% 
  # calculate dissimilarity etc within TFR intervals
  summarize(divergence = sum(abs(`high MAB` - `low MAB`)) / 2,
            similarity = 1 - divergence) %>% 
  # plot it; same conclusion as we had before!
  ggplot(aes(x = TFRint, y = similarity)) +
  geom_line() +
  labs(title = "Higher TFR implies more similar fertility age distributions")
```


