---
title: "Session notes day 5"
author: "Tim Riffe"
date: '2022-07-08'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in some new data

Read in our example data for today, which in the first instance will serve as a prop for demonstrating `if` and `for` concepts.
```{r}
library(tidyverse)
library(janitor)
hfd <- 
  read.table(file = "Data/asfrRR.txt", 
             header = TRUE, 
             skip = 2, 
             na.strings = ".", 
             as.is = TRUE) %>% 
  clean_names() %>% 
  mutate(age = parse_number(age)) 
```

## if

```{r}
a <- runif(10)
if_else(a > .5, .8, .2)
a > .5
if_else(a > .5, true = "A", false = "B")
```

## nested if-else, 
You see it on the street, but just say no

This works, but it's laborious to write, and also a bit hard to read.
We show it so you understand it when you see others do it. But don't do it.
```{r}
ifelse(a > .8,  .8,
        ifelse(a > .6,  .6,
                ifelse(a > .4,  .4, NA)))
```

Another:

Design a `case_when()` recode to go from specific to general conditions. Each test is comma separated; and there is a left (logical test) and right (result) separated by `~`. Be sure to account for ALL cases that might be in the data. The catch-all at the end is to force any **leftovers** to be `TRUE` and give them some value. If you don't account for the **leftovers** then you jsut get NAs back. If you explicitly account for things that need to be `NA`, then you need to match the `NA` type (e.g. `NA_real_`, and for this it's just best to use autocomplete).
```{r}
case_when(
  a > .8 ~ .8,
  a > .6 ~ .6,
  a > .4 ~ .4
  TRUE ~ 0.1
)
```
## recoding many things

To recode lots of values, then use a join technique. Construct a look-up table, then join it to get the new values.

to find that weird expression I googled:
"regex r remove inside parentheses"
regex is for detecting patterns in text. That's all you need know for this situation.
```{r}
lookup <- 
  read_csv("https://raw.githubusercontent.com/timriffe/BSSD2022Module1/master/Data/code_lookup.csv") %>% 
  # separate(code, into = c("code", NA), sep = " ")
 mutate(code = gsub(pattern = "\\s*\\([^\\)]+\\)", 
                    replacement = "", 
                    x = code),
        # remove that trailing space, using position tricks
        code = substr(code, 1, nchar(code) - 1))
```

Time for the join; here the join is premised on having a nice look-up table where codes conform with what you have in your data. 

1. `left_join()` is left-side dominant (you lose no cases from the left object)
2. `right_join()` is the opposite
3. `inner_join()` keeps only matching cases
4. `full_join()` throws nothing out
```{r}
nrow(hfd)
hfd %>% 
  right_join(lookup, by = "code")
```

## plainold `if`

Sometimes `A`, `B` and `D` will appear, other times not. It depends on whether `a[1]` is greater than .5, and it was randomly generated.

The structure of and `if` statement is as follows:
1. `if` starts it off
2. on the same line `(condition)` something that will evaluate to TRUE or FALSE. Just one value. You could call this a trigger.
3. `{ }` separates the code to be conditionally executed.
```{r}

if (a[1] > .5){

  A <- 25^a[1]
  B <- mean(a[-1])

  D <- A * B
}
```

What if we want to also have an alternate code chunk `{ }` inside another set of curly braces, then you can set this off with `else`. Further, these things can be nested to your heart's desire.

```{r}

# Calculate A, B and D if the value < .5
# first chunk
if (a[1] < .5){

  A <- 25^a[1]
  B <- mean(a[-1])

  D <- A * B
  if (D > 100){
    print("yay")
  } else {
    print("boo")
  }
 # first chunk ends here
} else { # what we do if value >= .5
  A <- sd(a)
} # ends the big if - else code 
```

```{r}

print("This code chunk was blocked off")
library(lubridate)

if (today() %>% weekdays() == "Friday"){
# some slow heavy code here


# slow heavy code ends here
}
```
## for -loops

A for-loop specifies an iterator, i.e. a value that changes in sequence, and being drawn from the elements of a vector. In this case, we're picking out letters, in order, from the Base R object called `letters`. The variable that changes in each iteration we called `letter` (singular). You can then use the iterator (or not) as you please within the loop. `{}` demarks the body of the code that is being iteratively executed.

```{r}
for (i in letters){
  plot(.5,.5, ylim = c(0,1), xlim = c(0,1), pch = i, cex = 10)
  Sys.sleep(.5)
  # repeat this code for each element of the iterator
}
```

Calculate the cumulative fertility rate for a population and year; Here the iterator is used to index positions in the vector that we need to calulate the result, as well as in the vector where we store the results. We sum the current value of `asfr` and the previous value of `casfr` in order to get the current value of `casfr`.
```{r}
BEL2015 <-
  hfd %>% 
  filter(code == "BEL",
         year == 2015)
asfr <- BEL2015$asfr
casfr <- asfr * NA
casfr[1] <- asfr[1]

for (age_i in 2:length(asfr)){
  casfr[age_i] <- casfr[age_i - 1] + asfr[age_i]
}
asfr[1:10]
for (i in 1:length(asfr)){
  casfr[i] <- sum(asfr[1:i])
}
# identical
# cumsum(asfr)
```

Tiny exercise. Calculate TFR for all HFD countries and years.

```{r}
tfr_looped <-
  hfd %>% 
  select(code, year) %>% 
  distinct() %>% 
  mutate(TFR = NA)

# i is a position index, we go down the rows of tfr_looped
for (i in 1:nrow(tfr_looped)){
  # select the current code and year
  this_code <- tfr_looped$code[i]
  this_year <- tfr_looped$year[i]
  
  # extract the data chunk we're at
  chunk     <- hfd %>% 
                 filter(code == this_code,
                        year == this_year)
  
  # calculate the TFR
  TFRi <- sum(chunk$asfr)
  
  # assign it to it's position in the container
  tfr_looped$TFR[i] <- TFRi
}

```
Let's contrast with the tidy approach to doing the same and look at pros and cons:

```{r}
tfr_tidy <-
  hfd %>% 
  group_by(code, year) %>% 
  summarize(TFR = sum(asfr),
            .groups = "drop")

tfr_tidy %>% View()
tfr_looped %>% View()
```

Pros of loops: 
1. it's easy to spot check what would happen at a particular iteration. So good for checking but also for investigating pathological cases.
2. non-`R` programmers can look at it and know what it's doing. It's a little more universal in nature and translatable.

Cons of loops:
1. they take a lot of set-up and possibly also planning
2. possible ambiguity about whether an iterator is being used to index a position (like us) or as an actual value
3. quite verbose and cluttery.

Pros of tidy approach in this case:
1. far more compact
2. easier to remember how to write it
3. easy to verbalize and wrap your mind around
4. more standard forms
5. easier to spot problems





## NOTE: funny encoding issue for Lucie, to be resolved later
```{r}
default_locale()
```




